\section{Reading assignment: John Stuart Mill -- On Liberty}

\subsection{Introduction}

\begin{itemize}
	\item concerned with Civil, or Social, Liberty: the nature and limits
	of power that can be legitimately exercised over an individual by the
	society
	\item ancient governments (except Greek democracy) always had an
	antagonism between the ruler and the ruled
	\item \textit{To prevent the weaker members of the community
	from being preyed upon by innumerable vultures, it
	was needful that there should be an animal of prey
	stronger than the rest, commissioned to keep them
	down.}
	\item but in reality the rulers usually preyed on the people
	\item the attempt to limit the rulers' power was first called liberty
	\item the attempts at liberty were done in two ways:
	\begin{itemize}
		\item by obtaining a recognition of certain political
		immunities, called political liberties or rights
		\item by establishment of constitutional checks
	\end{itemize}
	\item over time, people learned to prefer to delegate power to the
	government, not alienate it
	\item in the next step, people wanted the rulers to be identified with
	the people; that their will should be the interest of the nation
	\item protection against the tyranny of the magistrate is not enough.
	Harmful laws can be also passed by the public opinion.
	\item \textit{There is a limit to the legitimate
	interference of collective opinion with individual independence; and
	to find that limit, and maintain it against encroachment, is as
	indispensable to a good condition of human affairs, as protection
	against political despotism.}
\end{itemize}

The essay's main principle: \textit{that the sole end for which mankind are
warranted, individually or collectively in interfering with the liberty of
action of any of their number, is self-protection. That the only purpose for
which power can be rightfully exercised over any member of a civilized
community, against his will, is to prevent harm to others. His own good,
either physical or moral, is not a sufficient warrant.}

\section{Reading assignment: Hampton on utilitarianism}

Jeremy Bentham (1748-1832)

\textbf{Classical utilitarianism}: The objective is to maximize the sum of
happiness of all members in the community.

\textbf{Average utilitarianism}: The objective to maximize the average
happiness among community members.

Criticisms:

\begin{itemize}
	\item Takes for granted that everyone can evaluate their own happiness.
	\item Assumes that this evaluation can be made also by those at power
	in a state.
	\item Assumes that happiness is quantifiable.
	\item Assumes that a happiness of one person can be added to that of
	another and it works in a linear fashion.
\end{itemize}

Mill was a follower of utilitarianism but he also criticised Bentham:

\begin{itemize}
	\item Experiences of pleasure differ not only in quantity but also in
	quality.
	\item Disliked the idea that there is "no higher end than pleasure".
\end{itemize}

\textbf{Interpersonal comparison of utility problem}

\textbf{Diminishing marginal utility}

Preference satisfaction is not always a good measure for what human welfare
means for example what if one's preference is to rape a woman or steal a purse.
